### 2017-05-02 paper.tex
initial commit

\documentclass{article}

\usepackage{cmcphys}
\usepackage[margin=1.0in]{geometry}
\usepackage[titletoc,title]{appendix}
\allowdisplaybreaks

\renewcommand{\L}{\mathcal{L}}

\title{Chaotic Behavior of the Triple Pendulum \\
    \large A Computational Approach }
\author{Rachel Bass \and Cory McCartan}
\date{May 2017}

\begin{document}

    \maketitle

    \section{Introduction}

    \clearpage
    \begin{appendices}

    \section{Equations of Motion}

    \begin{wrapfigure}[13]{r}{0.25\textwidth}
        \centering
        \includegraphics[width=0.2\textwidth]{drawing.tikz}
        \caption{The triple pendulum.}
    \end{wrapfigure}

    We can describe the position of a given mass by its Cartesian coordinates
    $(x_i, y_i)$.  The position of a given mass is the sum of its position
    relative to its pivot point and the position of its pivot point:
    \begin{align}
        x_1 &= l\sin\phi_1     & y_1 &= l(1-\cos\phi_1) \\
        x_2 &= x_1+l\sin\phi_2 & y_2 &= y_1 + l(1-\cos\phi_2) \\
        x_3 &= x_2+l\sin\phi_3 & y_3 &= y_2 + l(1-\cos\phi_3).
    \end{align}
    Making the appropriate substitutions, we have
    \begin{align}
        x_1 &= l\sin\phi_1     \\
        y_1 &= l(1-\cos\phi_1) \\
        x_2 &= l(\sin\phi_1+\sin\phi_2)  \\
        y_2 &= l((1-\cos\phi_1)+(1-\cos\phi_2)) \\
        x_3 &= l(\sin\phi_1+\sin\phi_2+\sin\phi_3)  \\
        y_3 &= l((1-\cos\phi_1)+(1-\cos\phi_2)+(1-\cos\phi_3)).
    \end{align}
    Taking time derivatives to find velocity, we have
    \begin{align}
        \dot x_1 &= l\dot\phi_1\cos\phi_1 \\
        \dot y_1 &= l\dot\phi_1\sin\phi_1 \\
        \dot x_2 &= l(\dot\phi_1\cos\phi_1+\dot\phi_2\cos\phi_2) \\
        \dot y_2 &= l(\dot\phi_1\sin\phi_1+\dot\phi_2\sin\phi_2) \\
        \dot x_3 &= l(\dot\phi_1\cos\phi_1+\dot\phi_2\cos\phi_2
            +\dot\phi_3\cos\phi_3) \\
        \dot y_3 &= l(\dot\phi_1\sin\phi_1+\dot\phi_2\sin\phi_2
            +\dot\phi_3\sin\phi_3).
    \end{align}
    For the kinetic energy, we will need to find $v^2={\dot x}^2{\dot y}^2$ for
    each mass. 
    \begin{align}
        v_1^2 &= {\dot x_1}^2+{\dot y_1}^2 \\
        &= (l\dot\phi_1\cos\phi_1)^2
            +(l\dot\phi_1\sin\phi_1)^2 \\
        &= l^2(\dot\phi_1^2\cos^2\phi_1+\dot\phi_1^2\sin^2\phi_1) \\
        &= l^2\dot\phi_1^2 \\\\
                %
        v_2^2 &= {\dot x_2}^2+{\dot y_2}^2 \\
        &= (l(\dot\phi_1\cos\phi_1+\dot\phi_2\cos\phi_2))^2
          +(l(\dot\phi_1\sin\phi_1+\dot\phi_2\sin\phi_2))^2 \\
        &= l^2\left( \dot\phi_1^2\cos^2\phi_1 
            + 2\dot\phi_1\dot\phi_2\cos\phi_1\cos\phi_2 
            + \dot\phi_2^2\cos^2\phi_2 + \dot\phi_1^2\sin^2\phi_1 
            + 2\dot\phi_1\dot\phi_2\sin\phi_1\sin\phi_2 
            + \dot\phi_2^2\sin^2\phi_2 \right) \\
        &= l^2\left( \dot\phi_1^2\cos^2\phi_1 + \dot\phi_1^2\sin^2\phi_1
            + \dot\phi_2^2\cos^2\phi_2 + \dot\phi_2^2\sin^2\phi_2
            + 2\dot\phi_1\dot\phi_2(\cos\phi_1\cos\phi_2 + \sin\phi_1\sin\phi_2)
             \right) \\
        &= l^2\left( \dot\phi_1^2 + \dot\phi_2^2
            + 2\dot\phi_1\dot\phi_2\cos(\phi_1 - \phi_2) \right) \\\\
                %
        v_3^2 &= {\dot x_3}^2+{\dot y_3}^2 \\
        &= (l(\dot\phi_1\cos\phi_1+\dot\phi_2\cos\phi_2+\dot\phi_3\cos\phi_3))^2 
          +(l(\dot\phi_1\sin\phi_1+\dot\phi_2\sin\phi_2+\dot\phi_3\sin\phi_3))^2 \\
        &= l^2\bigl( \dot\phi_1^2\sin^2\phi_1+\dot\phi_2^2\sin^2\phi_2
           +\dot\phi_3^2\sin^2\phi_3 + 2\dot\phi_1\dot\phi_2\sin\phi_1\sin\phi_2
           +2\dot\phi_1\dot\phi_3\sin\phi_1\sin\phi_3
           +2\dot\phi_2\dot\phi_3\sin\phi_2\sin\phi_3 \\
           &\quad +\dot\phi_1^2\cos^2\phi_1+\dot\phi_2^2\cos^2\phi_2
           +\dot\phi_3^2\cos^2\phi_3 + 2\dot\phi_1\dot\phi_2\cos\phi_1\cos\phi_2
           +2\dot\phi_1\dot\phi_3\cos\phi_1\cos\phi_3
           +2\dot\phi_2\dot\phi_3\cos\phi_2\cos\phi_3
        \bigr) \\
        &= l^2\left( \dot\phi_1^2+\dot\phi_2^2 +\dot\phi_3^2 
           +2\dot\phi_1\dot\phi_2\cos(\phi_1-\phi_2)
           +2\dot\phi_1\dot\phi_3\cos(\phi_1-\phi_3)
           +2\dot\phi_2\dot\phi_3\cos(\phi_2-\phi_3) \right) 
    \end{align}
    We can now write the kinetic and potential energy for the system by summing
    up $-mgy_i$ and $\frac{1}{2}mv_i^2$ for each mass $i$. We will drop the
    constant terms from the potential energy expression since they will be lost
    when we take the partial derivatives of the Lagrangian:
    \begin{align}
        U &= U_1 + U_2 + U_3 \\
        &= -mgl\left(3\cos\phi_1+2\cos\phi_2+\cos\phi_3 \right) \\\\
                %
        T &= T_1 + T_2 + T_3 \\
        &= \frac{1}{2}ml^2 \left( 3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_3^2
           + 4\dot\phi_1\dot\phi_2\cos(\phi_1-\phi_2)
           + 2\dot\phi_1\dot\phi_3\cos(\phi_1-\phi_3)
           + 2\dot\phi_2\dot\phi_3\cos(\phi_2-\phi_3) \right) 
    \end{align}
    So the Lagrangian is
    \begin{align}
        \L &= T + U \\
        &= -mgl\left(3\cos\phi_1+2\cos\phi_2+\cos\phi_3 \right) 
           + \frac{1}{2}ml^2 \bigl( 3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_3^2
           + 4\dot\phi_1\dot\phi_2\cos(\phi_1-\phi_2) \\
        &\qquad\qquad + 2\dot\phi_1\dot\phi_3\cos(\phi_1-\phi_3)
           + 2\dot\phi_2\dot\phi_3\cos(\phi_2-\phi_3) \bigr) 
    \end{align}
    Since our generalized coordinates are $\phi_1,\phi_2$, and $\phi_3$, we must
    now take partial derivatives of the Lagrangian with respect to each $\phi_i$
    and $\dot\phi_i$:
    \begin{align}
        \pD{\L}{\phi_1} &= 3mgl\sin\phi_1 
            -2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
            -ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3) \\
        \pD{\L}{\phi_2} &= 2mgl\sin\phi_2 
            +2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
            -ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3) \\
        \pD{\L}{\phi_3} &= mgl\sin\phi_3 
            +ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
            +ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3) \\\\
                %
        \pD{\L}{\dot\phi_1} &= 3ml^2\dot\phi_1 
            + 2ml^2\dot\phi_2\cos(\phi_1-\phi_2) 
            + ml^2\dot\phi_3\cos(\phi_1-\phi_3)  \\
        \pD{\L}{\dot\phi_2} &= 2ml^2\dot\phi_2 
            + 2ml^2\dot\phi_1\cos(\phi_1-\phi_2) 
            + ml^2\dot\phi_3\cos(\phi_2-\phi_3)  \\
        \pD{\L}{\dot\phi_3} &= ml^2\dot\phi_3 
            + ml^2\dot\phi_1\cos(\phi_1-\phi_3) 
            + ml^2\dot\phi_2\cos(\phi_2-\phi_3) 
    \end{align}
    Next, we find $\D{}{t}\pD{\L}{\dot\phi_i}$ for each $i$:
    \begin{align}
        \D{}{t}\pD{\L}{\dot\phi_1} &= 3ml^2\ddot\phi_1 
            + 2ml^2\ddot\phi_2\cos(\phi_1-\phi_2) 
            - 2ml^2\dot\phi_2\sin(\phi_1-\phi_2)(\dot\phi_1-\dot\phi_2) \\
            &\quad + ml^2\ddot\phi_3\cos(\phi_1-\phi_3) 
            - ml^2\dot\phi_3\sin(\phi_1-\phi_3)(\dot\phi_1-\dot\phi_3) \\\\
        \D{}{t}\pD{\L}{\dot\phi_2} &= 2ml^2\ddot\phi_2 
            + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2) 
            - 2ml^2\dot\phi_1\sin(\phi_1-\phi_2)(\dot\phi_1-\dot\phi_2) \\
            &\quad + ml^2\ddot\phi_3\cos(\phi_2-\phi_3) 
            - ml^2\dot\phi_3\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) \\\\
        \D{}{t}\pD{\L}{\dot\phi_3} &= ml^2\ddot\phi_3 
            + ml^2\ddot\phi_1\cos(\phi_1-\phi_3) 
            - ml^2\dot\phi_1\sin(\phi_1-\phi_3)(\dot\phi_1-\dot\phi_3) \\
            &\quad + ml^2\ddot\phi_2\cos(\phi_2-\phi_3) 
            - ml^2\dot\phi_2\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) 
    \end{align}
    We can then write the equations of motion for each of our generalized
    coordinates, according to the Euler-Lagrange condition $\pD{\L}{\phi_i}
    =\D{}{t}\pD{\L}{\dot\phi_i}$. For $\phi_1$ we have
    \begin{align}
        \pD{\L}{\phi_1} &= \D{}{t}\pD{\L}{\dot\phi_1} 
    \end{align}
    \begin{align}
        3mgl\sin\phi_1 -2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
            -ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
        &= 3ml^2\ddot\phi_1 + 2ml^2\ddot\phi_2\cos(\phi_1-\phi_2) \\
        &\quad - 2ml^2\dot\phi_2\sin(\phi_1-\phi_2)(\dot\phi_1-\dot\phi_2) \\
        &\quad + ml^2\ddot\phi_3\cos(\phi_1-\phi_3) \\
        &\quad - ml^2\dot\phi_3\sin(\phi_1-\phi_3)(\dot\phi_1-\dot\phi_3) 
    \end{align}
    \begin{align}
        3mgl\sin\phi_1 &= 3ml^2\ddot\phi_1 + 2ml^2\ddot\phi_2\cos(\phi_1-\phi_2)
            + ml^2\ddot\phi_3\cos(\phi_1-\phi_3)  \\
            &\quad + 2ml^2\dot\phi_2^2\sin(\phi_1-\phi_2)
            - ml^2\dot\phi_3^2\sin(\phi_1-\phi_3)
    \end{align}
    \begin{align}
        \frac{3g}{l}\sin\phi_1 &= 3\ddot\phi_1 + 2\ddot\phi_2\cos(\phi_1-\phi_2)
            + \ddot\phi_3\cos(\phi_1-\phi_3) + 2\dot\phi_2^2\sin(\phi_1-\phi_2)
            - \dot\phi_3^2\sin(\phi_1-\phi_3)   \label{eq_m1}
    \end{align}

    For $\phi_2$ we have
    \begin{align}
        \pD{\L}{\phi_2} &= \D{}{t}\pD{\L}{\dot\phi_2} 
    \end{align}
    \begin{align}
        2mgl\sin\phi_2 +2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
            -ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
        &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2) \\
        &\quad - 2ml^2\dot\phi_1\sin(\phi_1-\phi_2)(\dot\phi_1-\dot\phi_2) \\
        &\quad + ml^2\ddot\phi_3\cos(\phi_2-\phi_3) \\
        &\quad - ml^2\dot\phi_3\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) 
    \end{align}
    \begin{align}
        2mgl\sin\phi_2 &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2)
            + ml^2\ddot\phi_3\cos(\phi_2-\phi_3)  \\
            &\quad + 2ml^2\dot\phi_1^2\sin(\phi_1-\phi_2)
            - ml^2\dot\phi_3^2\sin(\phi_2-\phi_3)
    \end{align}
    \begin{align}
        \frac{2g}{l}\sin\phi_2 &= 2\ddot\phi_2 + 2\ddot\phi_1\cos(\phi_1-\phi_2)
            + \ddot\phi_3\cos(\phi_2-\phi_3) + 2\dot\phi_1^2\sin(\phi_1-\phi_2)
            - \dot\phi_3^2\sin(\phi_2-\phi_3)   \label{eq_m2}
    \end{align}

    For $\phi_3$ we have
    \begin{align}
        \pD{\L}{\phi_3} &= \D{}{t}\pD{\L}{\dot\phi_3} 
    \end{align}
    \begin{align}
        mgl\sin\phi_3 +ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
            +ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
        &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3) \\
        &\quad - ml^2\dot\phi_1\sin(\phi_1-\phi_3)(\dot\phi_1-\dot\phi_3) \\
        &\quad + ml^2\ddot\phi_2\cos(\phi_2-\phi_3) \\
        &\quad - ml^2\dot\phi_2\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) 
    \end{align}
    \begin{align}
        mgl\sin\phi_3 &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3)
            + ml^2\ddot\phi_2\cos(\phi_2-\phi_3)  \\
            &\quad + ml^2\dot\phi_1^2\sin(\phi_1-\phi_3)
            - ml^2\dot\phi_2^2\sin(\phi_2-\phi_3)
    \end{align}
    \begin{align}
        \frac{g}{l}\sin\phi_3 &= \ddot\phi_3 + \ddot\phi_1\cos(\phi_1-\phi_3)
            + \ddot\phi_2\cos(\phi_2-\phi_3) + \dot\phi_1^2\sin(\phi_1-\phi_3)
            - \dot\phi_2^2\sin(\phi_2-\phi_3)   \label{eq_m3}
    \end{align}

    Rearranging $\eqref{eq_m1}$, $\eqref{eq_m2}$, and $\eqref{eq_m3}$ by moving
    the second derivatives to one side, we have
    \begin{align}
         3\ddot\phi_1 + 2\ddot\phi_2\cos(\phi_1-\phi_2) + \ddot\phi_3\cos(\phi_1-\phi_3) + 
             &= -2\dot\phi_2^2\sin(\phi_1-\phi_2) + \dot\phi_3^2\sin(\phi_1-\phi_3)
             +\frac{3g}{l}\sin\phi_1 \\
        2\ddot\phi_1\cos(\phi_1-\phi_2)+ 2\ddot\phi_2 + \ddot\phi_3\cos(\phi_2-\phi_3)
            &= - 2\dot\phi_1^2\sin(\phi_1-\phi_2) + \dot\phi_3^2\sin(\phi_2-\phi_3)
            +\frac{2g}{l}\sin\phi_2 \\
        \ddot\phi_1\cos(\phi_1-\phi_3+ \ddot\phi_2\cos(\phi_2-\phi_3)) + \ddot\phi_3
            &= - \dot\phi_1^2\sin(\phi_1-\phi_3) + \dot\phi_2^2\sin(\phi_2-\phi_3)
            +\frac{g}{l}\sin\phi_3
    \end{align}
    We can rewrite this system  equations as a single matrix equation:
    \begingroup 
    \begin{equation} 
        \renewcommand*{\arraystretch}{1.75} % more vertical space in matrix
        \resizebox{0.95\textwidth}{!}{$
        \begin{pmatrix}
            3                    & 2\cos(\phi_1-\phi_2) & \cos(\phi_1-\phi_3) \\
            2\cos(\phi_1-\phi_2) & 2                    & \cos(\phi_2-\phi_3) \\
            \cos(\phi_1-\phi_3)  & \cos(\phi_2-\phi_3)  & 1                   \\
        \end{pmatrix}
        \begin{pmatrix}
            \ddot\phi_1 \\
            \ddot\phi_2 \\
            \ddot\phi_3
        \end{pmatrix} =
        \begin{pmatrix}
            \dot\phi_3^2\sin(\phi_1-\phi_3) -2\dot\phi_2^2\sin(\phi_1-\phi_2)
                +\frac{3g}{l}\sin\phi_1 \\
            \dot\phi_3^2\sin(\phi_2-\phi_3) - 2\dot\phi_1^2\sin(\phi_1-\phi_2)
                +\frac{2g}{l}\sin\phi_2 \\
            \dot\phi_2^2\sin(\phi_2-\phi_3) -\dot\phi_1^2\sin(\phi_1-\phi_3)
                +\frac{g}{l}\sin\phi_3
        \end{pmatrix}
        $}
    \end{equation} 
    \endgroup 
    We can then solve for our vector of second derivatives by multiplying on the
    left by the inverse of the first matrix:
    \begingroup 
    \begin{equation} \renewcommand*{\arraystretch}{1.75} % more vertical space in matrix
        \resizebox{0.95\textwidth}{!}{$
        \begin{pmatrix}
            \ddot\phi_1 \\
            \ddot\phi_2 \\
            \ddot\phi_3
        \end{pmatrix} =
        \begin{pmatrix}
            3                    & 2\cos(\phi_1-\phi_2) & \cos(\phi_1-\phi_3) \\
            2\cos(\phi_1-\phi_2) & 2                    & \cos(\phi_2-\phi_3) \\
            \cos(\phi_1-\phi_3)  & \cos(\phi_2-\phi_3)  & 1                   \\
        \end{pmatrix}^{-1}
        \begin{pmatrix}
            \dot\phi_3^2\sin(\phi_1-\phi_3) -2\dot\phi_2^2\sin(\phi_1-\phi_2)
                +\frac{3g}{l}\sin\phi_1 \\
            \dot\phi_3^2\sin(\phi_2-\phi_3) - 2\dot\phi_1^2\sin(\phi_1-\phi_2)
                +\frac{2g}{l}\sin\phi_2 \\
            \dot\phi_2^2\sin(\phi_2-\phi_3) -\dot\phi_1^2\sin(\phi_1-\phi_3)
                +\frac{g}{l}\sin\phi_3
        \end{pmatrix} 
        $} \label{mat-eqn}
    \end{equation} 
    \endgroup 
    The matrix inverse in the above equation is very tedious to calculate. We
    used a computer algebra system\footnote{SageMath, the Sage Mathematics
    Software System (Version 7.6), The Sage Developers, 2017, http://www.sagemath.org.}
    to do the inverse and the matrix multiplication afterwards. The result is
    \begingroup 
    \begin{equation} 
        \renewcommand*{\arraystretch}{2.5} % more vertical space in matrix
        \resizebox{0.98\textwidth}{!}{$
        \begin{pmatrix}
            \ddot\phi_1 \\
            \ddot\phi_2 \\
            \ddot\phi_3
        \end{pmatrix} =
        \begin{pmatrix}
            \frac{1}{3} {\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{2 {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} + \frac{2 \cos(\phi_1-\phi_2)^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 1\right)}- {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} + \frac{\cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}\right)} + \frac{2 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} \\
            -{\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} + \frac{\cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}\right)} + \frac{3}{2} {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{1}{2 \cos(\phi_1-\phi_2)^2- 3} + \frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)}^2 {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}}\right)}-\frac{3 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)} {\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} \\
            \frac{2 {\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} + \frac{6 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)}}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6}-\frac{3 {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}}
        \end{pmatrix}
        $}\, ,\vspace{0.5em}
    \end{equation} 
    \endgroup 
    which is far too messy to use computationally. Instead, we chose to
    construct the matrices in $\eqref{mat-eqn}$ in \texttt{numpy} and
    numerically calculate the inverse and multiplication. Since many of the
    terms are repeated (especially the trigonometric functions of angle
    differences), this has the added advantage of allowing us to precompute
    these functions.

    \end{appendices}

\end{document}

### 2017-05-03 paper.tex
started ipynb
continuing work

@@
 \renewcommand{\L}{\mathcal{L}}
+\newcommand{\e}{\mathrm{e}}
+\renewcommand{\i}{\mathrm{i}}
+\renewcommand{\Re}{\mathrm{Re}\ }
 
@@
     \section{Equations of Motion}
+    \subsection{Full Solution}
 
@@
     used a computer algebra system\footnote{SageMath, the Sage Mathematics
-    Software System (Version 7.6), The Sage Developers, 2017, http://www.sagemath.org.}
+        Software System (Version 7.6), The Sage Developers, 2017,
+        \texttt{http://www.sagemath.org}.}
     to do the inverse and the matrix multiplication afterwards. The result is
@@
         \begin{pmatrix}
-            \frac{1}{3} {\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{2 {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} + \frac{2 \cos(\phi_1-\phi_2)^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 1\right)}- {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} + \frac{\cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}\right)} + \frac{2 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} \\
-            -{\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} + \frac{\cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}\right)} + \frac{3}{2} {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{1}{2 \cos(\phi_1-\phi_2)^2- 3} + \frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)}^2 {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}}\right)}-\frac{3 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)} {\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} \\
+            \frac{1}{3} {\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g \sin\left( \phi_1\right)}{l}\right)} {\left(\frac{2 {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} + \frac{2 \cos(\phi_1-\phi_2)^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 1\right)}- {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} + \frac{\cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}\right)} + \frac{2 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} \\
+            -{\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g\sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} + \frac{\cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}\right)} + \frac{3}{2} {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{1}{2 \cos(\phi_1-\phi_2)^2- 3} + \frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)}^2 {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}}\right)}-\frac{3 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)} {\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}} \\
            \frac{2 {\left(2 \dot\phi_2^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_1-\phi_3)-\frac{3 g \sin\left(\phi_1\right)}{l}\right)} {\left(\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)} \cos(\phi_1-\phi_2)}{2 \cos(\phi_1-\phi_2)^2- 3}-\cos(\phi_1-\phi_3)\right)}}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6} + \frac{6 {\left(\dot\phi_1^2 \sin(\phi_1-\phi_3)-\dot\phi_2^2 \sin(\phi_2-\phi_3)-\frac{g \sin\left(\phi_1\right)}{l}\right)}}{2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6}-\frac{3 {\left(2 \dot\phi_1^2 \sin(\phi_1-\phi_2)-\dot\phi_3^2 \sin(\phi_2-\phi_3)-\frac{2 g \sin\left(\phi_1\right)}{l}\right)} {\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}}{{\left(2 \cos(\phi_1-\phi_2)^2- 3\right)} {\left(2 \cos(\phi_1-\phi_3)^2-\frac{{\left(2 \cos(\phi_1-\phi_2) \cos(\phi_1-\phi_3)- 3 \cos(\phi_2-\phi_3)\right)}^2}{2 \cos(\phi_1-\phi_2)^2- 3}- 6\right)}}
@@
 
+    \subsection{Small-angle Approximation}
+
+    To check the validity of our numerical solution, we would like an analytical
+    solution. Clearly a full analytical solution is impossible. However, by
+    assuming that $\phi_1,\phi_2$, and $\phi_3$ remain small, we can radically
+    simplify the equations of motion to reach a closed-form expression for
+    $\phi_1,\phi_2$, and $phi_3$ as a function of time.
+
+    We begin by substituting $\cos\phi_i=1-\frac{1}{2}\phi_i^2$, the second
+    order Maclaurin approximation for cosine, into our expression for potential
+    energy. We also substitute $\dot\phi_i\dot\phi_j\cos(\phi_i-\phi_j)=\dot
+    \phi_i\dot\phi_j$, since for the small product $\dot\phi_i\dot\phi_j$, the
+    cosine term is negligible. This allows us to simplify our expression for
+    kinetic energy as well. We find
+    \begin{align}
+        U &= -mgl\left(3(1-\tfrac{1}{2}\phi_1^2)+2(1-\tfrac{1}{2}\phi_2^2
+            +(1-\tfrac{1}{2}\phi_3^2))\right) \\
+        &\simeq mgl\left(\frac{3}{2}\phi_1^2+\frac{2}{2}\phi_2^2
+            +\frac{1}{2}\phi_3^2\right) \tag*{(since the constant terms will be
+            dropped when the derivative is taken)} \\\\
+        T &= \frac{1}{2}ml^2(3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_1^2
+            +4\dot\phi_1\phi_2+2\dot\phi_1\phi_3+2\dot\phi_2\phi_3),
+    \end{align}
+    so the Lagrangian is
+    \begin{align}
+        \L &= T+U \\
+        &= \frac{1}{2}ml^2(3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_1^2
+            +4\dot\phi_1\phi_2+2\dot\phi_1\phi_3+2\dot\phi_2\phi_3)
+            -mgl\left(\frac{3}{2}\phi_1^2+\frac{2}{2}\phi_2^2
+            +\frac{1}{2}\phi_3^2\right)
+    \end{align}
+    Taking derivatives as required, we have
+    \begin{align}
+        \pD{\L}{\phi_1} &= -3mgl\phi_1, & 
+        \pD{\L}{\dot\phi_1} &= ml^2(3\dot\phi_1+4\dot\phi_2+2\dot\phi_3)
+        &\Rightarrow\quad
+        \D{}{t}\pD{\L}{\dot\phi_1} &= ml^2(3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3) \\
+                %
+        \pD{\L}{\phi_2} &= -2mgl\phi_2, & 
+        \pD{\L}{\dot\phi_2} &= ml^2(4\dot\phi_1+2\dot\phi_2+2\dot\phi_3)
+        &\Rightarrow\quad
+        \D{}{t}\pD{\L}{\dot\phi_3} &= ml^2(4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3) \\
+                %
+        \pD{\L}{\phi_3} &= -mgl\phi_3, & 
+        \pD{\L}{\dot\phi_3} &= ml^2(2\dot\phi_1+2\dot\phi_2+\dot\phi_3)
+        &\Rightarrow\quad
+        \D{}{t}\pD{\L}{\dot\phi_3} &= ml^2(2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3) \\
+    \end{align}
+    Applying our equilibrium condition $\D{}{t}\pD{\L}{\dot\phi_i}=\pD{\L}
+    {\phi_i}$, we have our three equations of motion:
+    \begin{align}
+        ml^2(3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3) &= -3mgl\phi_1 \\
+        ml^2(4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3) &= -2mgl\phi_2 \\
+        ml^2(2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3) &= -mgl\phi_3
+    \end{align}
+    Cancelling the common term of $ml^2$, we have
+    \begin{align}
+        3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3 &= -\frac{3g}{l}\phi_1 \\
+        4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3 &= -\frac{2g}{l}\phi_2 \\
+        2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3  &= -\frac{g}{l}\phi_3
+    \end{align}
+    We can write this system of equations as a single matrix equation \[
+        M\vec{\ddot\phi} = -G\vec{\phi} 
+    \] where
+    \begin{align}
+        M &= \begin{pmatrix}3&4&2\\4&2&2\\2&2&1\end{pmatrix} &
+        G &= \frac{g}{l}\begin{pmatrix}3&0&0\\0&2&0\\0&0&1\end{pmatrix}
+    \end{align}
+    We assume a periodic solution of the form \[
+        \vec\phi = \Re \vec{z}, \quad\mathrm{where}\quad
+        \vec{z} = \vec{a}\,\e^{\i\omega t}
+    \] Taking the second derivative of $\vec{z}$, we find \[
+        \vec{\ddot z} = -\omega^2\vec{a}\,\e^{\i\omega t} = -omega^2\vec{z}
+    \] So our differential equation becomes \[
+        -\omega^2 M\vec{a}\e^{\i\omega t} = -G\vec{a}\e^{\i\omega t},
+    \] which by rearranging and cancelling the common exponential term yields \[
+        (G-\omega^2M)\vec{a} = 0
+    \] This equation will have nontrivial solutions when $\det(G-\omega^2M)=0$.
+    Calculating the determinant (again with a computer algebra system), we find
+    \begin{align}
+        w^{6}+\frac{9g}{l}w^4+\frac{9g^2}{l^2}w^2-\frac{3g^3}{l^3} &= 0
+    \end{align}
+
     \end{appendices}

### 2017-05-09 paper.tex
further work

@@
     \section{Equations of Motion}
-    \subsection{Full Solution}
 
@@
     \begin{align}
-        x_1 &= l\sin\phi_1     & y_1 &= l(1-\cos\phi_1) \\
-        x_2 &= x_1+l\sin\phi_2 & y_2 &= y_1 + l(1-\cos\phi_2) \\
-        x_3 &= x_2+l\sin\phi_3 & y_3 &= y_2 + l(1-\cos\phi_3).
+        x_1 &= l\sin\phi_1     & y_1 &= -l\cos\phi_1 \\
+        x_2 &= x_1+l\sin\phi_2 & y_2 &= y_1 + -l\cos\phi_2 \\
+        x_3 &= x_2+l\sin\phi_3 & y_3 &= y_2 + -l\cos\phi_3.
     \end{align}
@@
         x_1 &= l\sin\phi_1     \\
-        y_1 &= l(1-\cos\phi_1) \\
+        y_1 &= -l\cos\phi_1 \\
         x_2 &= l(\sin\phi_1+\sin\phi_2)  \\
-        y_2 &= l((1-\cos\phi_1)+(1-\cos\phi_2)) \\
+        y_2 &= -l(\cos\phi_1+\cos\phi_2) \\
         x_3 &= l(\sin\phi_1+\sin\phi_2+\sin\phi_3)  \\
-        y_3 &= l((1-\cos\phi_1)+(1-\cos\phi_2)+(1-\cos\phi_3)).
+        y_3 &= -l(\cos\phi_1+\cos\phi_2+\cos\phi_3).
     \end{align}
@@
     We can now write the kinetic and potential energy for the system by summing
-    up $-mgy_i$ and $\frac{1}{2}mv_i^2$ for each mass $i$. We will drop the
-    constant terms from the potential energy expression since they will be lost
-    when we take the partial derivatives of the Lagrangian:
+    up $-mgy_i$ and $\frac{1}{2}mv_i^2$ for each mass $i$:
     \begin{align}
@@
     \begin{align}
-        \L &= T + U \\
-        &= -mgl\left(3\cos\phi_1+2\cos\phi_2+\cos\phi_3 \right) 
+        \L &= T - U \\
+        &= mgl\left(3\cos\phi_1+2\cos\phi_2+\cos\phi_3 \right) 
            + \frac{1}{2}ml^2 \bigl( 3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_3^2
@@
     \begin{align}
-        \pD{\L}{\phi_1} &= 3mgl\sin\phi_1 
+        \pD{\L}{\phi_1} &= -3mgl\sin\phi_1 
             -2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
             -ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3) \\
-        \pD{\L}{\phi_2} &= 2mgl\sin\phi_2 
+        \pD{\L}{\phi_2} &= -2mgl\sin\phi_2 
             +2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
             -ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3) \\
-        \pD{\L}{\phi_3} &= mgl\sin\phi_3 
+        \pD{\L}{\phi_3} &= -mgl\sin\phi_3 
             +ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
@@
     \begin{align}
-        3mgl\sin\phi_1 -2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
+        -3mgl\sin\phi_1 -2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
             -ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
@@
     \begin{align}
-        3mgl\sin\phi_1 &= 3ml^2\ddot\phi_1 + 2ml^2\ddot\phi_2\cos(\phi_1-\phi_2)
+        -3mgl\sin\phi_1 &= 3ml^2\ddot\phi_1 + 2ml^2\ddot\phi_2\cos(\phi_1-\phi_2)
             + ml^2\ddot\phi_3\cos(\phi_1-\phi_3)  \\
             &\quad + 2ml^2\dot\phi_2^2\sin(\phi_1-\phi_2)
-            - ml^2\dot\phi_3^2\sin(\phi_1-\phi_3)
+            + ml^2\dot\phi_3^2\sin(\phi_1-\phi_3)
     \end{align}
     \begin{align}
-        \frac{3g}{l}\sin\phi_1 &= 3\ddot\phi_1 + 2\ddot\phi_2\cos(\phi_1-\phi_2)
+        -\frac{3g}{l}\sin\phi_1 &= 3\ddot\phi_1 + 2\ddot\phi_2\cos(\phi_1-\phi_2)
             + \ddot\phi_3\cos(\phi_1-\phi_3) + 2\dot\phi_2^2\sin(\phi_1-\phi_2)
-            - \dot\phi_3^2\sin(\phi_1-\phi_3)   \label{eq_m1}
+            + \dot\phi_3^2\sin(\phi_1-\phi_3)   \label{eq_m1}
     \end{align}
@@
     \begin{align}
-        2mgl\sin\phi_2 +2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
+        -2mgl\sin\phi_2 +2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
             -ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
@@
     \begin{align}
-        2mgl\sin\phi_2 &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2)
+        -2mgl\sin\phi_2 &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2)
             + ml^2\ddot\phi_3\cos(\phi_2-\phi_3)  \\
-            &\quad + 2ml^2\dot\phi_1^2\sin(\phi_1-\phi_2)
-            - ml^2\dot\phi_3^2\sin(\phi_2-\phi_3)
+            &\quad - 2ml^2\dot\phi_1^2\sin(\phi_1-\phi_2)
+            + ml^2\dot\phi_3^2\sin(\phi_2-\phi_3)
     \end{align}
     \begin{align}
-        \frac{2g}{l}\sin\phi_2 &= 2\ddot\phi_2 + 2\ddot\phi_1\cos(\phi_1-\phi_2)
-            + \ddot\phi_3\cos(\phi_2-\phi_3) + 2\dot\phi_1^2\sin(\phi_1-\phi_2)
-            - \dot\phi_3^2\sin(\phi_2-\phi_3)   \label{eq_m2}
+        -\frac{2g}{l}\sin\phi_2 &= 2\ddot\phi_2 + 2\ddot\phi_1\cos(\phi_1-\phi_2)
+            + \ddot\phi_3\cos(\phi_2-\phi_3) - 2\dot\phi_1^2\sin(\phi_1-\phi_2)
+            + \dot\phi_3^2\sin(\phi_2-\phi_3)   \label{eq_m2}
     \end{align}
@@
     \begin{align}
-        mgl\sin\phi_3 +ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
+        -mgl\sin\phi_3 +ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
             +ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
@@
     \begin{align}
-        mgl\sin\phi_3 &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3)
+        -mgl\sin\phi_3 &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3)
             + ml^2\ddot\phi_2\cos(\phi_2-\phi_3)  \\
-            &\quad + ml^2\dot\phi_1^2\sin(\phi_1-\phi_3)
+            &\quad - ml^2\dot\phi_1^2\sin(\phi_1-\phi_3)
             - ml^2\dot\phi_2^2\sin(\phi_2-\phi_3)
@@
     \begin{align}
-        \frac{g}{l}\sin\phi_3 &= \ddot\phi_3 + \ddot\phi_1\cos(\phi_1-\phi_3)
-            + \ddot\phi_2\cos(\phi_2-\phi_3) + \dot\phi_1^2\sin(\phi_1-\phi_3)
+        -\frac{g}{l}\sin\phi_3 &= \ddot\phi_3 + \ddot\phi_1\cos(\phi_1-\phi_3)
+            + \ddot\phi_2\cos(\phi_2-\phi_3) - \dot\phi_1^2\sin(\phi_1-\phi_3)
             - \dot\phi_2^2\sin(\phi_2-\phi_3)   \label{eq_m3}
@@
          3\ddot\phi_1 + 2\ddot\phi_2\cos(\phi_1-\phi_2) + \ddot\phi_3\cos(\phi_1-\phi_3) + 
-             &= -2\dot\phi_2^2\sin(\phi_1-\phi_2) + \dot\phi_3^2\sin(\phi_1-\phi_3)
-             +\frac{3g}{l}\sin\phi_1 \\
+             &= -2\dot\phi_2^2\sin(\phi_1-\phi_2) - \dot\phi_3^2\sin(\phi_1-\phi_3)
+             -\frac{3g}{l}\sin\phi_1 \\
         2\ddot\phi_1\cos(\phi_1-\phi_2)+ 2\ddot\phi_2 + \ddot\phi_3\cos(\phi_2-\phi_3)
-            &= - 2\dot\phi_1^2\sin(\phi_1-\phi_2) + \dot\phi_3^2\sin(\phi_2-\phi_3)
-            +\frac{2g}{l}\sin\phi_2 \\
-        \ddot\phi_1\cos(\phi_1-\phi_3+ \ddot\phi_2\cos(\phi_2-\phi_3)) + \ddot\phi_3
-            &= - \dot\phi_1^2\sin(\phi_1-\phi_3) + \dot\phi_2^2\sin(\phi_2-\phi_3)
-            +\frac{g}{l}\sin\phi_3
+            &= 2\dot\phi_1^2\sin(\phi_1-\phi_2) - \dot\phi_3^2\sin(\phi_2-\phi_3)
+            -\frac{2g}{l}\sin\phi_2 \\
+        \ddot\phi_1\cos(\phi_1-\phi_3)+\ddot\phi_2\cos(\phi_2-\phi_3)+\ddot\phi_3
+            &= + \dot\phi_1^2\sin(\phi_1-\phi_3) + \dot\phi_2^2\sin(\phi_2-\phi_3)
+            -\frac{g}{l}\sin\phi_3
     \end{align}
@@
         \begin{pmatrix}
-            \dot\phi_3^2\sin(\phi_1-\phi_3) -2\dot\phi_2^2\sin(\phi_1-\phi_2)
-                +\frac{3g}{l}\sin\phi_1 \\
-            \dot\phi_3^2\sin(\phi_2-\phi_3) - 2\dot\phi_1^2\sin(\phi_1-\phi_2)
-                +\frac{2g}{l}\sin\phi_2 \\
-            \dot\phi_2^2\sin(\phi_2-\phi_3) -\dot\phi_1^2\sin(\phi_1-\phi_3)
-                +\frac{g}{l}\sin\phi_3
+            -\dot\phi_3^2\sin(\phi_1-\phi_3) -2\dot\phi_2^2\sin(\phi_1-\phi_2)
+                -\frac{3g}{l}\sin\phi_1 \\
+            -\dot\phi_3^2\sin(\phi_2-\phi_3) + 2\dot\phi_1^2\sin(\phi_1-\phi_2)
+                -\frac{2g}{l}\sin\phi_2 \\
+            \dot\phi_2^2\sin(\phi_2-\phi_3) + \dot\phi_1^2\sin(\phi_1-\phi_3)
+                -\frac{g}{l}\sin\phi_3
         \end{pmatrix}
@@
 
-    \subsection{Small-angle Approximation}
-
-    To check the validity of our numerical solution, we would like an analytical
-    solution. Clearly a full analytical solution is impossible. However, by
-    assuming that $\phi_1,\phi_2$, and $\phi_3$ remain small, we can radically
-    simplify the equations of motion to reach a closed-form expression for
-    $\phi_1,\phi_2$, and $phi_3$ as a function of time.
-
-    We begin by substituting $\cos\phi_i=1-\frac{1}{2}\phi_i^2$, the second
-    order Maclaurin approximation for cosine, into our expression for potential
-    energy. We also substitute $\dot\phi_i\dot\phi_j\cos(\phi_i-\phi_j)=\dot
-    \phi_i\dot\phi_j$, since for the small product $\dot\phi_i\dot\phi_j$, the
-    cosine term is negligible. This allows us to simplify our expression for
-    kinetic energy as well. We find
-    \begin{align}
-        U &= -mgl\left(3(1-\tfrac{1}{2}\phi_1^2)+2(1-\tfrac{1}{2}\phi_2^2
-            +(1-\tfrac{1}{2}\phi_3^2))\right) \\
-        &\simeq mgl\left(\frac{3}{2}\phi_1^2+\frac{2}{2}\phi_2^2
-            +\frac{1}{2}\phi_3^2\right) \tag*{(since the constant terms will be
-            dropped when the derivative is taken)} \\\\
-        T &= \frac{1}{2}ml^2(3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_1^2
-            +4\dot\phi_1\phi_2+2\dot\phi_1\phi_3+2\dot\phi_2\phi_3),
-    \end{align}
-    so the Lagrangian is
-    \begin{align}
-        \L &= T+U \\
-        &= \frac{1}{2}ml^2(3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_1^2
-            +4\dot\phi_1\phi_2+2\dot\phi_1\phi_3+2\dot\phi_2\phi_3)
-            -mgl\left(\frac{3}{2}\phi_1^2+\frac{2}{2}\phi_2^2
-            +\frac{1}{2}\phi_3^2\right)
-    \end{align}
-    Taking derivatives as required, we have
-    \begin{align}
-        \pD{\L}{\phi_1} &= -3mgl\phi_1, & 
-        \pD{\L}{\dot\phi_1} &= ml^2(3\dot\phi_1+4\dot\phi_2+2\dot\phi_3)
-        &\Rightarrow\quad
-        \D{}{t}\pD{\L}{\dot\phi_1} &= ml^2(3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3) \\
-                %
-        \pD{\L}{\phi_2} &= -2mgl\phi_2, & 
-        \pD{\L}{\dot\phi_2} &= ml^2(4\dot\phi_1+2\dot\phi_2+2\dot\phi_3)
-        &\Rightarrow\quad
-        \D{}{t}\pD{\L}{\dot\phi_3} &= ml^2(4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3) \\
-                %
-        \pD{\L}{\phi_3} &= -mgl\phi_3, & 
-        \pD{\L}{\dot\phi_3} &= ml^2(2\dot\phi_1+2\dot\phi_2+\dot\phi_3)
-        &\Rightarrow\quad
-        \D{}{t}\pD{\L}{\dot\phi_3} &= ml^2(2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3) \\
-    \end{align}
-    Applying our equilibrium condition $\D{}{t}\pD{\L}{\dot\phi_i}=\pD{\L}
-    {\phi_i}$, we have our three equations of motion:
-    \begin{align}
-        ml^2(3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3) &= -3mgl\phi_1 \\
-        ml^2(4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3) &= -2mgl\phi_2 \\
-        ml^2(2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3) &= -mgl\phi_3
-    \end{align}
-    Cancelling the common term of $ml^2$, we have
-    \begin{align}
-        3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3 &= -\frac{3g}{l}\phi_1 \\
-        4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3 &= -\frac{2g}{l}\phi_2 \\
-        2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3  &= -\frac{g}{l}\phi_3
-    \end{align}
-    We can write this system of equations as a single matrix equation \[
-        M\vec{\ddot\phi} = -G\vec{\phi} 
-    \] where
-    \begin{align}
-        M &= \begin{pmatrix}3&4&2\\4&2&2\\2&2&1\end{pmatrix} &
-        G &= \frac{g}{l}\begin{pmatrix}3&0&0\\0&2&0\\0&0&1\end{pmatrix}
-    \end{align}
-    We assume a periodic solution of the form \[
-        \vec\phi = \Re \vec{z}, \quad\mathrm{where}\quad
-        \vec{z} = \vec{a}\,\e^{\i\omega t}
-    \] Taking the second derivative of $\vec{z}$, we find \[
-        \vec{\ddot z} = -\omega^2\vec{a}\,\e^{\i\omega t} = -omega^2\vec{z}
-    \] So our differential equation becomes \[
-        -\omega^2 M\vec{a}\e^{\i\omega t} = -G\vec{a}\e^{\i\omega t},
-    \] which by rearranging and cancelling the common exponential term yields \[
-        (G-\omega^2M)\vec{a} = 0
-    \] This equation will have nontrivial solutions when $\det(G-\omega^2M)=0$.
-    Calculating the determinant (again with a computer algebra system), we find
-    \begin{align}
-        w^{6}+\frac{9g}{l}w^4+\frac{9g^2}{l^2}w^2-\frac{3g^3}{l^3} &= 0
-    \end{align}
-
     \end{appendices}

### 2017-05-11 paper.tex
progress!

@@
 \allowdisplaybreaks
+
+\graphicspath{{figures/}}
 
@@
 \section{Introduction}
+In this paper, we will report on our findings regarding the behavior and
+motion of an undamped triple pendulum made of three equal masses attached by
+three massless rods of equal length. We used the Lagrangian formulation to
+create equations of motion for the three masses, then examined the system using
+Python. We used Fourier transformations as a tool to help us determine if the
+pendulum's motion was chaotic or periodic. \textbf{Did we examine how changing
+the masses affects the motion or are we giving up on that?} Our findings helped
+us find some order about a system that initially appeared to be purely chaotic.
+
+\section{Conservation of Energy}
+After finding our equations of motion for the three masses, we first wanted to
+test that they were correct. Our plots of the masses' angles versus time seemed
+reasonable, but we wanted to confirm our suspicions before delving deeper into
+the problem. We plotted kinetic, potential and total energy versus time on the
+same set of axes and saw that mechanical energy of the system is conserved
+(Fig. \ref{fig:energy_cons}).
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=\textwidth]{energy_conservation}
+	\caption{Kinetic, potential, and total energy over time for $\dot\phi_3(0)=\SI{7}{s^{-1}}$.}
+	\label{fig:energy_cons}
+\end{figure}
+
+\section{Chaotic or Periodic?}
+As we first began to examine the system, we kept all initial angles and angular
+velocities save for the lowest hanging mass's at zero. For short time frames
+and an initial angular velocity of $\dot\phi_3=\SI{7}{s^{-1}}$, the motion of
+the system appears to be chaotic. However, increasing the time window to 1,000
+seconds showed something different. In Figure \ref{fig:moderate_time}, we can
+see that there appears to be some sort of regularity to the pendulum's motion.
+Looking at such a plot is not sufficient to declare periodic victory, so we
+needed to use some other tools to help us describe the system's motion.
+ \begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{moderate_velocity_time_sol}
+	\caption{$\phi_1,\phi_2$, and $\phi_3$ over time for $\dot\phi_3(0)=\SI{7}{s^{-1}}$}
+	\label{fig:moderate_time}
+\end{figure}
+
+\section{Results}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{bifurcation_1}
+	\caption{``Bifurcation Diagram" for $2\le\dot\phi_3(0)\le 20$}
+	\label{fig:bifurc_1}
+\end{figure}
 

### 2017-05-14 paper.tex
working

--- 2017-05-09 paper.tex
@@
 
+\graphicspath{{figures/}}
+
+\renewcommand{\figurename}{Fig.}
 \renewcommand{\L}{\mathcal{L}}
@@
 
+\begin{wrapfigure}[16]{R}{0.2\textwidth}
+	\centering
+	\includegraphics[width=0.17\textwidth]{drawing.tikz}
+    \caption{The triple pendulum.} 
+    \label{fig:pendulum}
+\end{wrapfigure}
+
 \section{Introduction}
+The triple pendulum is a classic example of a mechanical system that can be
+described simply yet behaves chaotically under a wide variety of initial
+conditions. Compared to the double pendulum, the addition of another mass (and
+accompanying coordinate $\phi_3$) adds significantly more complexity and
+non-linearity to the system dynamics. We study the conditions under which the
+triple pendulum is chaotic for the special case of equal lengths and masses of
+the pendula (see Fig. \ref{fig:pendulum}). We identify several critical
+values for initial conditions, beyond which the pendulum generally behaves in a
+chaotic manner.
+
+\section{Modeling}
+Before we model the triple pendulum, we need to find the system's 
+equations of motion.  The Lagrangian of the system is
+\begin{align}
+    \L &=  mgl\left(3\cos\phi_1+2\cos\phi_2+\cos\phi_3 \right) 
+	   + \frac{1}{2}ml^2 \bigl( 3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_3^2 \\
+	   &\qquad\qquad + 4\dot\phi_1\dot\phi_2\cos(\phi_1-\phi_2)   
+	   + 2\dot\phi_1\dot\phi_3\cos(\phi_1-\phi_3)
+	   + 2\dot\phi_2\dot\phi_3\cos(\phi_2-\phi_3) \bigr). 
+\end{align}
+For more details, see Appendix A. 
+We make the simplifying assumptions that all three masses
+are equal, and that the connecting rods are massless and of equal length.
+From the Lagrangian, we find our equations of motion for the three masses. 
+These equations are most easily expressed as a matrix equation:
+\begingroup 
+\begin{equation} \renewcommand*{\arraystretch}{1.25} 
+	\resizebox{0.95\textwidth}{!}{$
+	\begin{pmatrix}
+		\ddot\phi_1 \\
+		\ddot\phi_2 \\
+		\ddot\phi_3
+	\end{pmatrix} =
+	\begin{pmatrix}
+		3                    & 2\cos(\phi_1-\phi_2) & \cos(\phi_1-\phi_3) \\
+		2\cos(\phi_1-\phi_2) & 2                    & \cos(\phi_2-\phi_3) \\
+		\cos(\phi_1-\phi_3)  & \cos(\phi_2-\phi_3)  & 1                   \\
+	\end{pmatrix}^{-1}
+	\begin{pmatrix}
+		-\dot\phi_3^2\sin(\phi_1-\phi_3) -2\dot\phi_2^2\sin(\phi_1-\phi_2)
+			-\frac{3g}{l}\sin\phi_1 \\
+		-\dot\phi_3^2\sin(\phi_2-\phi_3) + 2\dot\phi_1^2\sin(\phi_1-\phi_2)
+			-\frac{2g}{l}\sin\phi_2 \\
+		\dot\phi_2^2\sin(\phi_2-\phi_3) + \dot\phi_1^2\sin(\phi_1-\phi_3)
+			-\frac{g}{l}\sin\phi_3
+	\end{pmatrix}
+	$} 
+\end{equation} 
+\endgroup 
+Note that this matrix equation includes a matrix inverse. Instead of 
+performing this tedious calculation by hand, we take advantage of the 
+Python library \texttt{numpy} to perform the calculation.
+
+\section{Validation}
+
+\begin{wrapfigure}[10]{R}{0.5\textwidth}
+	\centering
+	\includegraphics[width=0.5\textwidth]{energy_conservation}
+	\caption{Kinetic, potential, and total energy over time for $\dot\phi_3(0)=\SI{7}{s^{-1}}$.}
+	\label{fig:energy_cons}
+\end{wrapfigure}
+
+There is no analytical solution to above equations of motion.  The best 
+way to verify our numerical solution is to check that energy is conserved.
+Manually calculating the kinetic and potential energy from the numerical 
+solution for the case with moderate initial angular velocity, we confirm 
+that total energy is indeed conserved (Fig. \ref{fig:energy_cons}).
+
+We also confirmed that with no initial angular velocity or displacement, 
+the pendulum system remained motionless.
+
+\section{Analysis}
+
+\subsection{Fourier Transform}
+Given the complexity of the system, it can be difficult to determine whether 
+a given solution is periodic or not by merely looking at its time plot 
+(Fig. \ref{fig:moderate_time}) or  phase space plot (the fact that the phase 
+space is six-dimensional compounds these difficulties).
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.8\textwidth]{moderate_velocity_time_sol}
+	\caption{$\phi_1,\phi_2$, and $\phi_3$ over time for 
+	$\dot\phi_3(0)=\SI{7}{s^{-1}}$.}
+	\label{fig:moderate_time}
+\end{figure}
+
+A robust way to determine periodicity is to use the Fourier transform of the 
+angle data. Periodic solutions have strong peaks at nonzero frequency values,
+while chaotic solutions have no peaks other than a large singularity at zero 
+(Fig. \ref{fig:fft}).
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=0.49\textwidth]{fft_periodic}
+    \includegraphics[width=0.49\textwidth]{fft_chaotic}
+    \caption{The Fourier transform for $\phi_3(t)$.
+        Left: $\dot\phi_3(0) = \SI{7}{s^{-1}}$.
+        Right: $\dot\phi_3(0) = \SI{15}{s^{-1}}$. }
+    \label{fig:fft}
+\end{figure}
+
+\subsection{Bifurcation Diagrams and FFT Contour Plots}
+In chaotic systems such as the damped driven pendulum, there is a natural
+frequency to the system, namely, the driving frequency. One can then sample 
+the solution at time steps corresponding to this frequency to produce a
+bifurcation diagram.  Unfortunately, there is no such natural frequency for
+the triple pendulum system, which makes a traditional bifurcation diagram 
+impossible.  We can, however, create a plot that can convey similar 
+information. As in the bifurcation diagram, we simulate the system across 
+a range of initial conditions. We take the Fast Fourier Transform (FFT) of 
+each solution, and plot this as a column of pixels, where the logarithm of 
+the Fourier transform is mapped to a color (magenta for low intensity, yellow 
+for high intensity). 
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=\textwidth]{contour_3}
+	\caption{FFT contour plot for $\SI{2}{s^{-1}}\le\dot\phi_3(0)
+    	\le\SI{20}{s^{-1}}$.}
+	\label{fig:contour_3}
+\end{figure}
+This allows us to see how the frequency composition of the
+solution changes as an initial parameter is changed---and, importantly, when 
+chaos sets in. We call this plot an FFT contour plot. Fig. \ref{fig:contour_3} 
+shows the FFT contour plot for $\phi_3$ as $\dot\phi_3(0)$ is varied. We can
+clearly see the two main frequency components from Fig. \ref{fig:fft} for 
+low values of $\dot\phi_3(0)$, but these abruptly disappear around $\dot
+\phi_3(0)=\SI{10.5}{s^{-1}}$. While there are short ``bursts" of periodicity
+past this point, for the most part, chaos dominates.
+
+\section{Results}
+
+\subsection{Non-chaotic Behavior}
+As seen in Fig. \ref{fig:contour_3}, the system is not chaotic for values of
+$\dot\phi_3(0)\lesssim\SI{10.5}{s^{-1}}$. Fig. \ref{fig:moderate_time} shows 
+the various angles over time, and Fig. \ref{fig:fft} shows on the left the 
+Fourier transform of $\phi_3$. We can also change coordinates from $\phi_1,
+\phi_2$, and $\phi_3$ to Cartesian coordinates, which allows us to plot
+the actual trajectories the masses trace out over time (Fig. 
+\ref{fig:moderate_trajectories}).
+
+\begin{wrapfigure}[13]{r}{0.5\textwidth}
+	\centering
+	\includegraphics[width=0.5\textwidth]{moderate_velocity_trajectory}
+	\caption{Mass trajectories for $\dot\phi_3(0)=\SI{7}{s^{-1}}$.}
+	\label{fig:moderate_trajectories}
+\end{wrapfigure}
+
+We see that the masses are confined to a fairly narrow region of space, and 
+are essentially swinging back and forth together (this is confirmed by examining
+the FFT plot for the non-chaotic region (Fig. \ref{fig:fft}), where the main
+frequency components for each angle are equal).
+
+We can also create phase space plots for the system 
+(Fig. \ref{fig:moderate_phase}).
+Because we have three generalized coordinates, the phase space is 
+six-dimensional. We plot the angle and angular velocity of each mass on 
+its own two-dimensional phase space plot, essentially looking at three 
+different slices of phase space.  As a result, the phase space trajectories 
+cross frequently. Notice that the phase space diagrams are symmetrical.
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{moderate_velocity_phase_space}
+	\caption{Phase space slices for $\dot\phi_3(0)=\SI{7}{s^{-1}}$.}
+	\label{fig:moderate_phase}
+\end{figure}
+
+\subsection{Finding Chaos}
+We can use FFT contour plots to identify chaotic regions for various 
+configurations of initial conditions. If instead of giving the third 
+pendulum some initial velocity, we start all of the masses at rest, but 
+at an equal angle from vertical, we get a much different FFT contour plot 
+(Fig. \ref{fig:contour_angles}). Here we see that there is periodic behavior 
+when $\phi_1(0) = \phi_2(0) = \phi_3(0) \leq 1.4$ (about $78.8\degree$). 
+All initial angles past this point result in chaotic behavior. We can 
+compare the periodic behavior seen in Fig. \ref{fig:moderate_trajectories} 
+and Fig. \ref{fig:moderate_phase} with the chaotic behaviors seen in Fig.
+\ref{fig:postcritical_trajectories} and Fig. \ref{fig:postcritical_phase}. 
+The motion is no longer periodic, and the phase space trajectories no 
+longer symmetrical.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{contour_angles}
+	\caption{Phase space slices for $0\le\phi_i(0)\le\frac{3\pi}{4}$.}
+	\label{fig:contour_angles}
+\end{figure}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{postcritical_angle_trajectory}
+    \caption{Mass trajectories for $\phi_1(0)=\frac{\pi}{3}+0.04, 
+    \phi_2(0)=\phi_3(0)=\frac{\pi}{3}$.}
+    \label{fig:postcritical_trajectories}
+\end{figure}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{postcritical_angle_phase_space}
+    \caption{Phase space slices for $\phi_1(0)=\frac{\pi}{3}+0.04, \phi_2(0) = 
+    \phi_3(0)=\frac{\pi}{3}$.}
+    \label{fig:postcritical_phase}
+\end{figure}
+
+What is special about $78.8\degree$? Examining Fig. 
+\ref{fig:postcritical_trajectories}, we notice that mass $m_1$ approaches, 
+but never quite crosses, horizontal.  Indeed, when the simulation is run 
+again with a slightly higher initial angle, $m_1$ crosses horizontal and 
+the motion becomes chaotic.
+
+Fig. \ref{fig:max_angle} plots the maximum angle reached by $m_1$ over an
+entire simulation versus the initial angles of the pendula. Below the
+critical angle (in the non-chaotic region), the maximum value of $\phi_1$
+increases approximately linearly as the initial angle increases. At the
+critical angle the maximum value of $\phi_1$ is $90\degree$. Past this
+point, the behavior is chaotic. So it seems that crossing $90\degree$ 
+causes chaotic behavior.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{max_angle_critical}
+    \caption{Plot of maximum value of $\phi_1$ versus the starting angles
+    of each $\m_i$. A vertical line is drawn at the critical angle of 
+    $78.8\degree$, and a horizontal line is drawn at $\phi_1=90\degree$.}
+    \label{fig:max_angle}
+\end{figure}
 
